Statistical Reporting Inconsistencies in Experimental Linguistics

### 2024-07-25 manuscript/manuscript.qmd
minor fixes

---
title: "Statistical Reporting Inconsistencies in Experimental Linguistics"
shorttitle: "Statistical Inconsistencies"
author:
  - name: Dara Leonard Jenssen Etemady
    corresponding: false
    email: daraetemady@gmail.com
    affiliations:
      - name: University of Oslo
        department: Department of Linguistics & Scandinavian Studies
        city: Oslo
        country: Norway
  - name: Timo B. Roettger
    corresponding: true
    orcid: 0000-0000-0000-0001
    email: timo.roettger@gmail.com
    affiliations:
      - name: University of Oslo
        department: Department of Linguistics & Scandinavian Studies
        city: Oslo
        country: Norway
author-note:
  status-changes: 
    affiliation-change: null
    deceased: null
  disclosures:
    study-registration: null
    data-sharing: null
    related-report: null
    conflict-of-interest: The authors have no conflict of interest to declare.
    financial-support: null
    gratitude: null
    authorship-agreements: Conceptionalization, Methodology, Validation, Formal Analysis, Review & Editing of Manuscript, Data Curation - T.B.R. & D.L.J.E.; Software, Investigation - D.L.J.E.; Writing of Original Draft, Visualization, Supervision - T.B.R.
abstract: "The present paper investigates the prevalence of statistical reporting consistencies across articles in eight experimental linguistics journals, published between 2000 and 2023. Using the R package “Statcheck”, we retrieved 39532 statistical test and assessed whether p-values were inconsistent with their test statistic and degrees of freedom. Half of all articles that used null hypothesis signficance testing contained at least one inconsistent p-value. One in eight articles contained an inconsistency that may have affected the statistical conclusion. The inconsistency rates were stable across journals and publication years. We offer actionable steps for authors, reviewers, and editors to remmedy this state-of-affairs."
keywords: [statistics, statcheck, reproducibility]
bibliography: bibliography.bib
format:
  apaquarto-docx: default
  apaquarto-html: default
  apaquarto-pdf:
    # Can be jou (journal), man (manuscript), stu (student), or doc (document)
    documentmode: doc
    keep-tex: true
header-includes: \usepackage{annotate-equations}
execute:
  echo: false
editor: 
  markdown: 
    wrap: sentence
---

# 1. Introduction
What we know about human language and its cognitive underpinnings is often informed by experimental data.
Researchers test theoretical predictions with these data using statistical tests.
Depending on the results of such tests, researchers make claims for or against theoretical assumptions.\
Since they play such an integral part in the reasoning process of experimentalists, both the data these tests are based on and the computational procedure of the tests itself should be transparent in order for other researchers to critically evaluate them.
Moreover things can go wrong.
To err is human.
Transparent sharing allows others to detect and correct human error.
In the recent years, the quantitative sciences have seen repeated calls to become more transparent and reproducible by sharing data and statistical protocols [@roettger2019researcher; @arvan2022reproducibility; @laurinavichyute2022share].
Despite these calls, sharing of statistical protocols is still rather rare across the language sciences [@bochynska2023reproducible].
If statistical procedures cannot be critically evaluated, human errors might be left undetected and thus remain uncorrected in the publication record.
And if undetected errors affect the decision procedure of the analysis, i.e. whether a hypothesis is rejected or accepted, these errors might lead to -at best- overconfident, -at worst- false theoretical conclusions.
The present paper will present evidence that the published literature in experimental linguistics contains a concerning amount of such statistical errors, a state of affairs which warrants more rigorous data sharing practices.

# 2. Statistal reporting inconsistencies
The null-hypothesis significance testing (NHST) framework is, to date, the most dominant statistical framework that researchers use to test hypotheses in the language sciences [@sondereggera2024advancements].
NHST tests are commonly reported in specific formats which usually contain the name of the test (e.g. F, t, χ2), a test statistic, the degrees of freedom of that test (if applicable), and the p-value, representing the probability of observing the data (or more extreme data) given the null hypothesis (i.e. given that the test statistic is zero) (see example 1):
&nbsp;

&nbsp;

>>>>

---
title: "Statistical Reporting Inconsistencies in Experimental Linguistics"
shorttitle: "Statistical Inconsistencies"
author:
  - name: Dara Leonard Jenssen Etemady
    corresponding: true
    email: daraetemady@gmail.com
    affiliations:
      - name: University of Oslo
        department: Department of Linguistics & Scandinavian Studies
        city: Oslo
        country: Norway
  - name: Timo B. Roettger
    corresponding: false
    orcid: 0000-0000-0000-0001
    email: timo.roettger@gmail.com
    affiliations:
      - name: University of Oslo
        department: Department of Linguistics & Scandinavian Studies
        city: Oslo
        country: Norway
author-note:
  status-changes: 
    affiliation-change: null
    deceased: null
  disclosures:
    study-registration: null
    data-sharing: null
    related-report: null
    conflict-of-interest: The authors have no conflict of interest to declare.
    financial-support: null
    gratitude: null
    authorship-agreements: Conceptionalization, Methodology, Validation, Formal Analysis, Review & Editing of Manuscript, Data Curation - T.B.R. & D.L.J.E.; Software, Investigation - D.L.J.E.; Writing of Original Draft, Visualization, Supervision - T.B.R.
abstract: "The present paper investigates the prevalence of statistical reporting consistencies across articles in eight experimental linguistics journals published between 2000 and 2023. Using the R package “Statcheck”, we retrieved 39532 statistical test and assessed whether p-values were consistent with their test statistic and degrees of freedom. Half of all articles that used null hypothesis significance testing contained at least one inconsistent p-value. One in eight articles contained an inconsistency that may have affected the statistical conclusion. The inconsistency rates were comparable across journals and seem stable over publication years. We discuss possible reasons for this high rate and offer actionable steps for authors, reviewers, and editors to remedy this state-of-affairs."
keywords: [statistics, statcheck, reproducibility, null hypothesis significance testing]
bibliography: bibliography.bib
format:
  apaquarto-docx: default
  apaquarto-html: default
  apaquarto-pdf:
    # Can be jou (journal), man (manuscript), stu (student), or doc (document)
    documentmode: doc
    keep-tex: true
header-includes: \usepackage{annotate-equations}
execute:
  echo: false
editor: 
  markdown: 
    wrap: sentence
---

# 1. Introduction

What we know about human language and its cognitive underpinnings is often informed by experimental data.
Researchers test theoretical predictions with their data using statistical tests.
Depending on the results of their tests, researchers make claims for or against theoretical assumptions.\
Since these tests play such an integral part in the argumentation process of experimentalists, both the data these tests are based on and the computational procedure of the tests itself should be transparent in order for other researchers to critically evaluate them.
But more than that, things can go wrong.
Humans are fallible; they make mistakes.
Transparent sharing allows others to detect and correct human error.
In the recent years, quantitative linguistics have seen repeated calls to become more transparent and reproducible by sharing data and statistical protocols [@roettger2019researcher; @arvan2022reproducibility; @laurinavichyute2022share].
Despite these calls, sharing of statistical protocols is still rather rare across the language sciences [@bochynska2023reproducible].
If statistical procedures cannot be critically evaluated, human errors might be left undetected and thus remain uncorrected in the publication record.
And if undetected errors affect the decision procedure of the analysis, i.e. whether a hypothesis is rejected or accepted, these errors might lead to -at best- overconfident, -at worst- false theoretical conclusions.
The present paper will present evidence that the published literature in experimental linguistics contains a concerning amount of such statistical errors, a state of affairs which warrants more rigorous data sharing practices.

# 2. Statistal reporting inconsistencies

The null-hypothesis significance testing (NHST) framework is, to date, the most dominant statistical framework that researchers use to test hypotheses in the language sciences [@sondereggera2024advancements].
NHST tests are commonly reported in specific formats which usually contain the name of the test (e.g. F, t, χ2), a test statistic, the degrees of freedom of that test (if applicable), and the p-value, representing the probability of observing the data (or more extreme data) given the null hypothesis (i.e. given that the test statistic is zero) (see example 1):

### 2024-07-25 manuscript/manuscript_anonymized.qmd
minor fixes

---
title: "Statistical Reporting Inconsistencies in Experimental Linguistics"
shorttitle: "Statistical Inconsistencies"
author:
    - name: anonymous
      affiliations:
      - name: anonymous
  # - name: Dara Leonard Jenssen Etemady
  #   corresponding: true
  #   email: daraetemady@gmail.com
  #   affiliations:
  #     - name: University of Oslo
  #       department: Department of Linguistics & Scandinavian Studies
  #       city: Oslo
  #       country: Norway
  # - name: Timo B. Roettger
  #   corresponding: false
  #   orcid: 0000-0000-0000-0001
  #   email: timo.roettger@gmail.com
  #   affiliations:
  #     - name: University of Oslo
  #       department: Department of Linguistics & Scandinavian Studies
  #       city: Oslo
  #       country: Norway
author-note:
  status-changes: 
    affiliation-change: null
    deceased: null
  disclosures:
    study-registration: null
    data-sharing: null
    related-report: null
    conflict-of-interest: The authors have no conflict of interest to declare.
    financial-support: null
    gratitude: null
    # authorship-agreements: Conceptionalization, Methodology, Validation, Formal Analysis, Review & Editing of Manuscript, Data Curation - T.B.R. & D.L.J.E.; Software, Investigation - D.L.J.E.; Writing of Original Draft, Visualization, Supervision - T.B.R.
abstract: "The present paper investigates the prevalence of statistical reporting consistencies across articles in eight experimental linguistics journals published between 2000 and 2023. Using the R package “Statcheck”, we retrieved 39532 statistical test and assessed whether p-values were consistent with their test statistic and degrees of freedom. Half of all articles that used null hypothesis signficance testing contained at least one inconsistent p-value. One in eight articles contained an inconsistency that may have affected the statistical conclusion. The inconsistency rates were comparable across journals and seem stable over publication years. We discuss possible reasons for this high rate and offer actionable steps for authors, reviewers, and editors to remedy this state-of-affairs."
keywords: [statistics, statcheck, reproducibility, null hypothesis significance testing]
bibliography: bibliography.bib
format:
  apaquarto-docx: default
  apaquarto-html: default
  apaquarto-pdf:
    # Can be jou (journal), man (manuscript), stu (student), or doc (document)
    documentmode: doc
    keep-tex: true
header-includes: \usepackage{annotate-equations}
execute:
  echo: false
editor: 
  markdown: 
    wrap: sentence
---

@@

All quantitative analyses were conducted using @rmanual. All derived data and corresponding R scripts are publicly available here: [https://github.com/timo-b-roettger/ErrorsLing](https://github.com/timo-b-roettger/ErrorsLing)

>>>>

---
title: "Statistical Reporting Inconsistencies in Experimental Linguistics"
shorttitle: "Statistical Inconsistencies"
author:
    - name: anonymous
      affiliations:
      - name: anonymous
  # - name: Dara Leonard Jenssen Etemady
  #   corresponding: true
  #   email: daraetemady@gmail.com
  #   affiliations:
  #     - name: University of Oslo
  #       department: Department of Linguistics & Scandinavian Studies
  #       city: Oslo
  #       country: Norway
  # - name: Timo B. Roettger
  #   corresponding: false
  #   orcid: 0000-0000-0000-0001
  #   email: timo.roettger@gmail.com
  #   affiliations:
  #     - name: University of Oslo
  #       department: Department of Linguistics & Scandinavian Studies
  #       city: Oslo
  #       country: Norway
author-note:
  status-changes: 
    affiliation-change: null
    deceased: null
  disclosures:
    study-registration: null
    data-sharing: null
    related-report: null
    conflict-of-interest: The authors have no conflict of interest to declare.
    financial-support: null
    gratitude: null
    # authorship-agreements: Conceptionalization, Methodology, Validation, Formal Analysis, Review & Editing of Manuscript, Data Curation - T.B.R. & D.L.J.E.; Software, Investigation - D.L.J.E.; Writing of Original Draft, Visualization, Supervision - T.B.R.
abstract: "The present paper investigates the prevalence of statistical reporting consistencies across articles in eight experimental linguistics journals published between 2000 and 2023. Using the R package “Statcheck”, we retrieved 39532 statistical test and assessed whether p-values were consistent with their test statistic and degrees of freedom. Half of all articles that used null hypothesis significance testing contained at least one inconsistent p-value. One in eight articles contained an inconsistency that may have affected the statistical conclusion. The inconsistency rates were comparable across journals and seem stable over publication years. We discuss possible reasons for this high rate and offer actionable steps for authors, reviewers, and editors to remedy this state-of-affairs."
keywords: [statistics, statcheck, reproducibility, null hypothesis significance testing]
bibliography: bibliography.bib
format:
  apaquarto-docx: default
  apaquarto-html: default
  apaquarto-pdf:
    # Can be jou (journal), man (manuscript), stu (student), or doc (document)
    documentmode: doc
    keep-tex: true
header-includes: \usepackage{annotate-equations}
execute:
  echo: false
editor: 
  markdown: 
    wrap: sentence
---

@@

All quantitative analyses were conducted using @rmanual. All derived data and corresponding R scripts are available [here](https://osf.io/gx3ub/?view_only=25957e06b5f449d6820b529eb0ae1937).
